\chapter{Acknowledgements}
I would first like to thank my thesis supervisors: Prof. Ferdinando Ametrano and Prof. Daniele Marazzina. The first for sharing with me his deep passion for the argument. He fueled my hunger of knowledge with his lessons and his willingness in helping people made my working period a lot easier. The latter for his valuable advice during the drawing up of the thesis.

\bigskip
\noindent
Then I would like to thank the project team I worked with during my intern period: Dr. Paolo Mazzocchi and Dr. Federico Giacomini, who taught me a lot of programming technicalities.

\bigskip
\noindent
A special thanks goes to my family, always by my side even in the most difficult situations of my life. To my mother, for loving me unconditionally. To my father, for believing in my intelligence more than I did myself. To my brother, for being an example. To my grandmother, for teaching me to make the best of things.  

\bigskip
\noindent
Moreover, I would like to deeply thank my friends Riccardo, Serena and Matteo. You have always taken care of me, urging me to never give up. Without your precious support, I would never have made it.

\bigskip
\noindent
Furthermore, I would like to thank all the friends who accompany me in everyday life: Piro, Spino, Santa, Peder, Ale, Molis, Guasta, Cri, Giulia, Silvia, Furi, Longo, Dada, Marti and all the others. You have always encouraged me to get involved in every situation, making me a better person.

\bigskip
\noindent
I would also like to thank my volleyball team. Making healthy sport is the best safety valve.

\bigskip
\noindent
Last, but absolutely not least, I would like to thank the Gellify team. You welcomed me with open arms, allowing me to work in a stress-free environment.