\chapter{Conclusions and Future Work}
\label{chpr:conclusions}
The aim of this thesis has been to give an explanation on how to strengthen the notarization process by means of a decentralized solution built upon a blockchain technology, in particular the Bitcoin blockchain, being the most secure and reliable decentralized system.

\bigskip
\noindent
To achieve this goal we first presented distributed system in general, emphasizing the advantages in using them and focusing on the consensus problem. Then we studied the consensus problem in presence of byzantine nodes in the case of an asynchronous network, putting ourselves in the same basic structure of Bitcoin. After presenting a famous result of Fisher, Lynch and Paterson \cite{Fischer:1985:IDC:3149.214121}, which proves the impossibility of distributed consensus with one faulty process, we saw how brilliantly Satoshi Nakamoto solves the consensus problem in Bitcoin, combining proof-of-work and economic incentives, thus making the underlying blockchain almost immutable. Hence, convinced of the security and immutability of the Bitcoin blockchain, we started presenting one of the most important non monetary applications of such technology, that is timestamping of digital documents. First we learned how the timestamping procedure on blockchain works, underlining the cryptographic operations needed. Then we introduced \textit{OpenTimestamps}, an open standard consisting in a set of operations for creating provable blockchain timestamps that can be independently verified. It also solves a scalability problem encountered in the first notarization attempts on blockchain, aggregating digests to be timestamped in a merkle tree. We also gave a step-by-step guide on the usage of the \textit{OpenTimestamps} Python client and the web interface, enticing the reader to try timestamping a document. Finally we developed a fully operating timestamping service, giving a detailed description of the solution, both in terms of IT architecture and underlying code, dividing the client side from the server side for a better understanding. Such solution marginally improves and extends \textit{OpenTimestamps} with additional features: multi-validation, Bitcoin testnet as an alternative chain and the check for the originality of a document separated from the verification process.

\bigskip
\noindent
An improvement of the timestamping process could be made thanks to elliptic curve commitments, which give the possibility to timestamp at zero marginal cost. Therefore, transaction fees can be avoided by pushing arbitrary data with specific commitment operations into the payee public key or in a digital signature. These techniques are described in the master thesis by Comandini \cite{Comandini:Thesis:2018}. A further improvement could be a more precise temporal resolution in the notarization process, no longer linked to the temporal precision in which a block is mined, the latter having a confidence interval of about two hours.
